---
title: "CD73 OverExpression in OV90 and OVCAR cell lines"
author: "Fabio Bedin | MS-Unit"
output: html_document
---


```{r librerie, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(fig.align = "center", warning=FALSE, message=FALSE, fig.height=8, fig.width=8)
library("MBQN")
library("dplyr")
library("tidyr")
library("DEP")
library("SummarizedExperiment")
library("preprocessCore")
library("tibble")
library("ggplot2")
library("enrichR")
library("DT")
library("stringr")
library("patchwork")
library("here")
library("datapasta")
library("ggstatsplot")
library("UpSetR")
library("ggrepel")
library("visdat")
library("naniar")
library("gprofiler2")
library("openxlsx")
library("org.Hs.eg.db")
library("rrvgo")
select <- dplyr::select
```

```{r Custom Functions}
source(here::here("code/custom_functions.R"))
```

```{r data}
data <- read.csv(here::here("data/proteinGroups_UC_CD73.txt"), header = TRUE,stringsAsFactors = FALSE, sep = "\t")

data <- data[data$Reverse != "+" & data$Potential.contaminant != "+" & data$Only.identified.by.site != "+" & data$Peptides >= 2,]

data_unique <- make_unique(data, "Gene.names", "Protein.IDs", delim = ";")

```

# **Introduction**

In this analysis we compare **OVCAR** and **OV90** cell lines, with or without overexpression of **CD73**.

```{r expdes}
expdesign <- read.table(here::here("data/experimentalDesignTemplate.txt"), header = T, stringsAsFactors = F)

columns<-match(paste("LFQ.intensity.",expdesign$label,sep=""),colnames(data_unique))

data_se <- make_se(data_unique, columns, expdesign)

data_filt <- filter_missval(data_se, thr = 1)
```

Fist we generate some informative prols:

## **Proteins per samples**

This plot show the number of proteins identify in each samples:

```{r proteins, fig.height=12, fig.width=15}
plot_numbers_lables(data_se, b_size = 18, lab_size = 7)
```

## **Upset plot** {.tabset .tabset-fade}

With this plot we can identify the numper of prteins in common within the 2 cell lines:

### OVCAR

```{r upset_1, fig.height=12, fig.width=15}
define_set <- assay(data_filt) %>%  as.data.frame() %>% select(starts_with("OVCAR")) %>% colnames()

assay(data_filt) %>% 
  as.data.frame() %>% 
  select(starts_with("OVCAR")) %>% 
  rownames_to_column() %>% 
  pivot_longer(!rowname, names_to = "samples", values_to = "intensity") %>% 
  mutate(intensity = if_else(is.na(intensity), 0, 1)) %>% 
  pivot_wider(names_from = samples, values_from = intensity) %>% 
  as.data.frame() %>% 
  upset(nsets = 8,
        sets = define_set,
        #order.by = "degree",
        order.by = "freq",
        keep.order = T,
        text.scale = 2.5,
        point.size = 4,
        line.size = 0.5, 
        sets.bar.color = rep(c("coral2", "turquoise3"), each = 4),
        main.bar.color  = "gray44")
  
```

### OV90

```{r upset_2, fig.height=12, fig.width=15}
define_set <- assay(data_filt) %>%  as.data.frame() %>% select(starts_with("OV90")) %>% colnames()

assay(data_filt) %>% 
  as.data.frame() %>% 
  select(starts_with("OV90")) %>% 
  rownames_to_column() %>% 
  pivot_longer(!rowname, names_to = "samples", values_to = "intensity") %>% 
  mutate(intensity = if_else(is.na(intensity), 0, 1)) %>% 
  pivot_wider(names_from = samples, values_from = intensity) %>% 
  as.data.frame() %>% 
  upset(nsets = 8,
        sets = define_set,
        #order.by = "degree",
        order.by = "freq",
        keep.order = T,
        text.scale = 2.5,
        point.size = 4,
        line.size = 0.5, 
        sets.bar.color = rep(c("coral2", "turquoise3"), each = 4),
        main.bar.color  = "gray44")
  
```

### Both cell lines

```{r upset_3, fig.height=12, fig.width=15}
define_set <- assay(data_filt) %>% colnames()

assay(data_filt) %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  pivot_longer(!rowname, names_to = "samples", values_to = "intensity") %>% 
  mutate(intensity = if_else(is.na(intensity), 0, 1)) %>% 
  pivot_wider(names_from = samples, values_from = intensity) %>% 
  as.data.frame() %>% 
  upset(nsets = 16,
        sets = define_set,
        #order.by = "degree",
        order.by = "freq",
        keep.order = T,
        text.scale = 2.5,
        point.size = 4,
        line.size = 0.5, 
        sets.bar.color = rep(c("salmon1", "red3", "turquoise3", "royalblue1"), each = 4),
        main.bar.color  = "gray44")
  
```

## **Differential enrichment analysis** {.tabset .tabset-fade}

Since we have 4 replicates per condition, we decided to filter the data for at least 3 out of 4 valid values in each condition.

```{r diff_2}
test <- c("OV90_CD73_vs_OV90_WT", "OVCAR_CD73_vs_OVCAR_WT")

data_diff <- test_diff_BH(data_filt, type = "manual", test = test)

dep <- add_rejections(data_diff, alpha = 0.05, lfc = 1)

results<-get_results(dep)
```

### PCA plot

```{r PCA_2}
plot_pca(dep, x = 1, y = 2, n = 500, point_size = 4, indicate = "condition")
```

### Volcano plot {.tabset .tabset-fade}

#### OVCAR

```{r ovcar_analysis}
expdesign <- read.table(here::here("data/experimentalDesignTemplate.txt"), header = T, stringsAsFactors = F)

conditions_CAR<-c("OVCAR_WT", "OVCAR_CD73")

expdesign <- subset(expdesign, condition %in% conditions_CAR)

columns<-match(paste("LFQ.intensity.",expdesign$label,sep=""),colnames(data_unique))

data_se_CAR <- make_se(data_unique, columns, expdesign)

data_filt <- filter_by_grouping(data_se_CAR, thr = 1)

test_CAR <- "OVCAR_CD73_vs_OVCAR_WT"

data_diff <- test_diff_BH(data_filt, type = "manual", test = test_CAR)

dep <- add_rejections(data_diff, alpha = 0.05, lfc = 1)

results_CAR<-get_results(dep)

plot_volcano(dep, contrast=test_CAR, add_names=T,label_size=5,adjusted = F)

results_DEP_OVCAR <- get_df_wide(dep) %>% 
  select(c(1:10,  Protein.names, starts_with("OVCAR"), -ends_with(c("CI.R", "CI.L")))) %>% 
  relocate(name, Protein.IDs, Protein.names) 
```

#### OV90

```{r ov90_analysis}
expdesign <- read.table(here::here("data/experimentalDesignTemplate.txt"), header = T, stringsAsFactors = F)

conditions_90<-c("OV90_WT", "OV90_CD73")

expdesign <- subset(expdesign, condition %in% conditions_90)

columns<-match(paste("LFQ.intensity.",expdesign$label,sep=""),colnames(data_unique))

data_se_90 <- make_se(data_unique, columns, expdesign)

data_filt <- filter_by_grouping(data_se_90, thr = 1)

test_90 <- "OV90_CD73_vs_OV90_WT"

data_diff <- test_diff_BH(data_filt, type = "manual", test = test_90)

dep <- add_rejections(data_diff, alpha = 0.05, lfc = 1)

results_90<-get_results(dep)

plot_volcano(dep, contrast=test_90, add_names=T,label_size=5,adjusted = F)

results_DEP_OV90 <- get_df_wide(dep) %>% 
  select(c(1:10,  Protein.names, starts_with("OV90"), -ends_with(c("CI.R", "CI.L")))) %>% 
  relocate(name, Protein.IDs, Protein.names)
```

### Results table {.tabset .tabset-fade}

In this interactive table are summarized the restults:

#### OVCAR

```{r result_car}
results_CAR %>% 
  select(starts_with("name") | starts_with(test_CAR)) %>% 
  filter(across(ends_with("significant"))) %>% 
  arrange(across(ends_with("p.adj"))) %>%
  mutate(across(2:3, format, scientific = T, digits = 2)) %>%
  dplyr::rename_with(~ tolower(gsub(paste0(test_CAR,"_"), "", .x)), .cols = starts_with(test_CAR)) %>% 
  mutate(significant = str_to_title(significant)) %>%
  dplyr::rename(FC = ratio) %>% 
  DT::datatable(options = list(
  columnDefs = list(list(className = 'dt-center', targets = 1:5)),
  pageLength = 10)) %>%
  formatStyle('FC', backgroundColor = styleInterval(c(-1, 1), c('lightblue', 'white', 'tomato'))) %>% 
  formatStyle('significant', color = styleEqual(c("True", "False"), c('green', 'red')))
```

#### OV90

```{r result_90}
results_90 %>% 
  select(starts_with("name") | starts_with(test_90)) %>% 
  filter(across(ends_with("significant"))) %>% 
  arrange(across(ends_with("p.adj"))) %>%
  mutate(across(2:3, format, scientific = T, digits = 2)) %>%
  dplyr::rename_with(~ tolower(gsub(paste0(test_90,"_"), "", .x)), .cols = starts_with(test_90)) %>% 
  mutate(significant = str_to_title(significant)) %>%
  dplyr::rename(FC = ratio) %>% 
  DT::datatable(options = list(
  columnDefs = list(list(className = 'dt-center', targets = 1:5)),
  pageLength = 10)) %>%
  formatStyle('FC', backgroundColor = styleInterval(c(-1, 1), c('lightblue', 'white', 'tomato'))) %>% 
  formatStyle('significant', color = styleEqual(c("True", "False"), c('green', 'red')))
```

#### Common DEP

```{r table_common}
results <- merge(results_CAR, results_90, by.x="ID", by.y="ID", all.x = TRUE, all.y = TRUE)

results %>% 
  mutate(name = case_when(is.na(name.x) ~ name.y, TRUE ~ name.x)) %>% 
  mutate(significant.x = case_when(is.na(significant.x) ~ FALSE, TRUE ~ significant.x)) %>% 
  mutate(significant.y = case_when(is.na(significant.y) ~ FALSE, TRUE ~ significant.y)) %>% 
  mutate(significant = if_else(significant.x & significant.y, TRUE, FALSE)) %>% 
  filter(significant) %>%
  select(-significant, -ends_with("centered"), -contains(c(".x", ".y"))) %>% 
  pivot_longer(
   cols = ends_with(c("p.val", "p.adj", "significant", "ratio")),
   names_to = c("comparison", ".value"),
   names_pattern = "(.*_.*_.*_.*_.*)_(.*)") %>% 
  arrange(name, p.adj) %>% 
  relocate(comparison, name) %>% 
  mutate(across(ends_with(c("p.val", "p.adj")), format, scientific = T, digits = 2)) %>% 
  mutate(significant = str_to_title(significant)) %>%
  dplyr::rename(FC = ratio) %>% 
  drop_na() %>%
  DT::datatable(options = list(
  columnDefs = list(list(className = 'dt-center', targets = 1:7)),
  pageLength = 10)) %>%
  formatStyle('FC', backgroundColor = styleInterval(c(-1, 1), c('lightblue', 'white', 'tomato'))) %>% 
  formatStyle('significant', color = styleEqual(c("True", "False"), c('green', 'red')))
```

## **Unique Proteins** {.tabset .tabset-fade}

### OVCAR

```{r unique_car}
data_filt_CAR <- filter_missval(data_se_CAR, thr = 1)

unique_1 <- assay(data_filt_CAR) %>% 
  as.data.frame() %>% 
  select(starts_with(conditions_CAR[1])) %>%
  rownames_to_column() %>% 
  pivot_longer(!rowname, names_to = "samples", values_to = "intensity") %>% 
  mutate(intensity = case_when(is.na(intensity) ~ 0, TRUE ~ as.numeric(1))) %>%
  group_by(rowname) %>% 
  mutate(unique = if_else(sum(intensity) == 0, T, F)) %>% 
  ungroup() %>% 
  filter(unique) %>% 
  pull(rowname) %>%
  unique()
  
unique_2 <- assay(data_filt_CAR) %>% 
  as.data.frame() %>% 
  select(starts_with(conditions_CAR[2])) %>%
  rownames_to_column() %>% 
  pivot_longer(!rowname, names_to = "samples", values_to = "intensity") %>% 
  mutate(intensity = case_when(is.na(intensity) ~ 0, TRUE ~ as.numeric(1))) %>%
  group_by(rowname) %>% 
  mutate(unique = if_else(sum(intensity) == 0, T, F)) %>% 
  ungroup() %>% 
  filter(unique) %>% 
  pull(rowname) %>%
  unique()

table_unique_CAR <- assay(data_filt_CAR) %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  mutate(!!paste0("unique_", conditions_CAR[2]) := rowname %in% unique_1) %>% 
  mutate(!!paste0("unique_", conditions_CAR[1]) := rowname %in% unique_2) %>% 
  filter(get(paste0("unique_", conditions_CAR[1])) | get(paste0("unique_", conditions_CAR[2]))) %>% 
  relocate(where(is.logical), .after = where(is.character)) %>% 
  dplyr::rename(name = rowname)

table_unique_CAR %>% 
  select(1:3) %>%
  DT::datatable(options = list(
  columnDefs = list(list(className = 'dt-center', targets = 1:3)),
  pageLength = 10)) %>%
  formatStyle('unique_OVCAR_CD73', color = styleEqual(c(TRUE, FALSE), c('green', 'red'))) %>% 
  formatStyle('unique_OVCAR_WT', color = styleEqual(c(TRUE, FALSE), c('green', 'red')))

result_unique_OVCAR <- get_df_wide(data_filt_CAR) %>% 
  select(c(name, Protein.IDs, Protein.names)) %>% 
  right_join(table_unique_CAR) 
```

### OV90

```{r unique_90}
data_filt_90 <- filter_missval(data_se_90, thr = 1)

unique_1 <- assay(data_filt_90) %>% 
  as.data.frame() %>% 
  select(starts_with(conditions_90[1])) %>%
  rownames_to_column() %>% 
  pivot_longer(!rowname, names_to = "samples", values_to = "intensity") %>% 
  mutate(intensity = case_when(is.na(intensity) ~ 0, TRUE ~ as.numeric(1))) %>%
  group_by(rowname) %>% 
  mutate(unique = if_else(sum(intensity) == 0, T, F)) %>% 
  ungroup() %>% 
  filter(unique) %>% 
  pull(rowname) %>%
  unique()
  
unique_2 <- assay(data_filt_90) %>% 
  as.data.frame() %>% 
  select(starts_with(conditions_90[2])) %>%
  rownames_to_column() %>% 
  pivot_longer(!rowname, names_to = "samples", values_to = "intensity") %>% 
  mutate(intensity = case_when(is.na(intensity) ~ 0, TRUE ~ as.numeric(1))) %>%
  group_by(rowname) %>% 
  mutate(unique = if_else(sum(intensity) == 0, T, F)) %>% 
  ungroup() %>% 
  filter(unique) %>% 
  pull(rowname) %>%
  unique()

table_unique_90 <- assay(data_filt_90) %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  mutate(!!paste0("unique_", conditions_90[2]) := rowname %in% unique_1) %>% 
  mutate(!!paste0("unique_", conditions_90[1]) := rowname %in% unique_2) %>% 
  filter(get(paste0("unique_", conditions_90[1])) | get(paste0("unique_", conditions_90[2]))) %>% 
  relocate(where(is.logical), .after = where(is.character)) %>% 
  dplyr::rename(name = rowname)

table_unique_90 %>% 
  select(1:3) %>%
  DT::datatable(options = list(
  columnDefs = list(list(className = 'dt-center', targets = 1:3)),
  pageLength = 10)) %>%
  formatStyle('unique_OV90_CD73', color = styleEqual(c(TRUE, FALSE), c('green', 'red'))) %>% 
  formatStyle('unique_OV90_WT', color = styleEqual(c(TRUE, FALSE), c('green', 'red')))

result_unique_OV90 <- get_df_wide(data_filt_90) %>% 
  select(c(name, Protein.IDs, Protein.names)) %>% 
  right_join(table_unique_90)
```


## **Gene Ontology** {.tabset .tabset-fade}

Then we perfom a gene onology analysis (GO or by branch GO:MF, GO:BP, GO:CC) and a KEGG ontology.

The enrichment results are visualized with an interactive **Manhattan-like-plot**:

### OVCAR {.tabset .tabset-fade}

#### CD73 

```{r GO_car_cd73}
Gname <- results_CAR %>% filter(significant & get(paste0(test_CAR, "_ratio")) > 0) %>% pull(name)

Uni <- table_unique_CAR %>% filter(unique_OVCAR_CD73) %>% pull(name) %>% unique()

#bg_list <- results_CAR %>% pull(name) %>% unique()

GO_name <- c(Gname, Uni)

GO <- gost(query = GO_name, organism = "hsapiens", sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG"), ordered_query = T)

gostplot(GO, capped = TRUE, interactive = TRUE)
```

#### Table CD73

The results can also be visualized with a table: 

```{r tab_car_cd73}
GO$result %>% 
  as.data.frame() %>% 
  select(starts_with(c("source", "term_name", "p_value", "term_size", "intersection_size"))) %>% 
  arrange(p_value) %>%  
  mutate(p_value = format(p_value, scientific = T, digits = 2)) %>% 
  DT::datatable(options = list(
  columnDefs = list(list(className = 'dt-center', targets = 1:6)),
  pageLength = 10)) %>%
  formatStyle('source', backgroundColor = styleEqual(c("GO:BP", "GO:MF", "GO:CC", "KEGG"), c('orange', 'tomato', 'lightgreen', 'violet')))
```

#### Reduced Terms CD73

```{r}
simMatrix <- GO$result %>% 
  as.data.frame() %>%
  filter(source == "GO:BP") %>% 
  pull(term_id) %>% 
  calculateSimMatrix(orgdb="org.Hs.eg.db", ont="BP", method="Rel")

scores <- setNames(-log10(GO$result$p_value), GO$result$term_id)

reducedTerms <- reduceSimMatrix(simMatrix,
                                scores,
                                threshold=0.7,
                                orgdb="org.Hs.eg.db")

treemapPlot(reducedTerms)
```


#### WT

```{r GO_car_wt}
Gname <- results_CAR %>% filter(significant & get(paste0(test_CAR, "_ratio")) < 0) %>% pull(name)

Uni <- table_unique_CAR %>% filter(unique_OVCAR_WT) %>% pull(name)

#bg_list <- results_CAR %>% pull(name) %>% unique()

GO_name <- c(Gname, Uni)

GO <- gost(query = GO_name, organism = "hsapiens", sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG"), ordered_query = T)

gostplot(GO, capped = TRUE, interactive = TRUE)
```

#### table WT

The results can also be visualized with a table:

```{r tab_car_wt}
GO$result %>% 
  as.data.frame() %>% 
  select(starts_with(c("source", "term_name", "p_value", "term_size", "intersection_size"))) %>% 
  arrange(p_value) %>%  
  mutate(p_value = format(p_value, scientific = T, digits = 2)) %>% 
  DT::datatable(options = list(
  columnDefs = list(list(className = 'dt-center', targets = 1:6)),
  pageLength = 10)) %>%
  formatStyle('source', backgroundColor = styleEqual(c("GO:BP", "GO:MF", "GO:CC", "KEGG"), c('orange', 'tomato', 'lightgreen', 'violet')))
```

#### Reduced Terms WT

```{r}
simMatrix <- GO$result %>% 
  as.data.frame() %>%
  filter(source == "GO:BP") %>% 
  pull(term_id) %>% 
  calculateSimMatrix(orgdb="org.Hs.eg.db", ont="BP", method="Rel")

scores <- setNames(-log10(GO$result$p_value), GO$result$term_id)

reducedTerms <- reduceSimMatrix(simMatrix,
                                scores,
                                threshold=0.7,
                                orgdb="org.Hs.eg.db")

treemapPlot(reducedTerms)
```



### OV90 {.tabset .tabset-fade}

#### CD73 

```{r GO_90_cd73}
Gname <- results_90 %>% filter(significant & get(paste0(test_90, "_ratio")) > 0) %>% pull(name)

Uni <- table_unique_90 %>% filter(unique_OV90_CD73) %>% pull(name)

#bg_list <- results_90 %>% pull(name) %>% unique()

GO_name <- c(Gname, Uni)

GO <- gost(query = GO_name, organism = "hsapiens", sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG"), ordered_query = T)

gostplot(GO, capped = TRUE, interactive = TRUE)
```

The results can also be visualized with a table: 

```{r tab_90_cd73}
GO$result %>% 
  as.data.frame() %>% 
  select(starts_with(c("source", "term_name", "p_value", "term_size", "intersection_size"))) %>% 
  arrange(p_value) %>%  
  mutate(p_value = format(p_value, scientific = T, digits = 2)) %>% 
  DT::datatable(options = list(
  columnDefs = list(list(className = 'dt-center', targets = 1:6)),
  pageLength = 10)) %>%
  formatStyle('source', backgroundColor = styleEqual(c("GO:BP", "GO:MF", "GO:CC", "KEGG"), c('orange', 'tomato', 'lightgreen', 'violet')))
```

#### WT

```{r GO_90_wt}
Gname <- results_90 %>% filter(significant & get(paste0(test_90, "_ratio")) < 0) %>% pull(name)

Uni <- table_unique_90 %>% filter(unique_OV90_WT) %>% pull(name)

#bg_list <- results_90 %>% pull(name) %>% unique()

GO_name <- c(Gname, Uni)

GO <- gost(query = GO_name, organism = "hsapiens", sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG"), ordered_query = T)

gostplot(GO, capped = TRUE, interactive = TRUE)
```

The results can also be visualized with a table:

```{r tab_90_wt}
GO$result %>% 
  as.data.frame() %>% 
  select(starts_with(c("source", "term_name", "p_value", "term_size", "intersection_size"))) %>% 
  arrange(p_value) %>%  
  mutate(p_value = format(p_value, scientific = T, digits = 2)) %>% 
  DT::datatable(options = list(
  columnDefs = list(list(className = 'dt-center', targets = 1:6)),
  pageLength = 10)) %>%
  formatStyle('source', backgroundColor = styleEqual(c("GO:BP", "GO:MF", "GO:CC", "KEGG"), c('orange', 'tomato', 'lightgreen', 'violet')))
```


## {-}


```{r}
#final_table <- get_df_wide(dep)
#write.table(final_table, file = "results_table.txt", quote = F, sep = "\t", row.names = F, col.names = T)

excel <- createWorkbook()

addWorksheet(excel, sheetName = "ProteinGroups_OVCAR")

writeDataTable(excel, sheet = "ProteinGroups_OVCAR", x = results_DEP_OVCAR)

addWorksheet(excel, sheetName = "Unique_OVCAR")

writeDataTable(excel, sheet = "Unique_OVCAR", x = result_unique_OVCAR)

addWorksheet(excel, sheetName = "ProteinGroups_OV90")

writeDataTable(excel, sheet = "ProteinGroups_OV90", x = results_DEP_OV90)

addWorksheet(excel, sheetName = "Unique_OV90")

writeDataTable(excel, sheet = "Unique_OV90", x = result_unique_OV90)

saveWorkbook(excel, here::here("output/results_UC_CD73_210709.xlsx"), overwrite = T)
```

